---
title: "Data Loading & Distribution: Historical Basketball Scores"
author: "Pushpendu Biswas"
date: today
format:
  html:
    theme:
      light: flatly
      dark: darkly
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

## Setup & Data Loading

This document loads historical basketball game records using **R** and the **tidyverse** library.

```{r load-data}
library(tidyverse)
library(plotly)

# Determine the exact filepath on disk (handling case-sensitivity)
file_path <- if (file.exists("Games.csv")) {
  "Games.csv"
} else if (file.exists("games.csv")) {
  "games.csv"
} else {
  stop("Could not find Games.csv or games.csv in the directory.")
}

# Read the data
games <- read_csv(file_path, show_col_types = FALSE)

# Display a preview of the loaded dataset
glimpse(games)
```

## Win Location Infographic

Below is an infographic showing the distribution of team wins by the type of arena (Home Arena or Away Arena).

```{r win-analysis}
# Categorize the win location
classified_games <- games %>%
  mutate(
    # A game is neutral if it's an All-Star game, OR if we have arena city data and it doesn't match home or away city
    is_neutral = (gameType == 'All-Star Game') | 
                 (!is.na(arenaCity) & !is.na(hometeamCity) & !is.na(awayteamCity) & 
                  arenaCity != hometeamCity & arenaCity != awayteamCity),
    
    win_location = case_when(
      is_neutral ~ 'Neutral Arena',
      winner == hometeamId ~ 'Home Arena',
      winner == awayteamId ~ 'Away Arena',
      TRUE ~ 'Unknown'
    )
  )

# Calculate win counts
counts <- classified_games %>% 
  count(win_location)

home_wins <- counts %>% filter(win_location == "Home Arena") %>% pull(n)
away_wins <- counts %>% filter(win_location == "Away Arena") %>% pull(n)
neutral_wins <- counts %>% filter(win_location == "Neutral Arena") %>% pull(n)
```

<div style="display: flex; flex-wrap: wrap; gap: 30px; align-items: stretch; margin: 30px 0;">
<div style="flex: 1; min-width: 280px; display: flex; flex-direction: column;">

```{r render-cards-html}
#| results: asis
# Calculate percentages based on home + away only
total_wins <- home_wins + away_wins
home_pct <- (home_wins / total_wins) * 100
away_pct <- (away_wins / total_wins) * 100

html_cards <- sprintf('
<div style="display: flex; flex-direction: column; gap: 20px; font-family: system-ui, -apple-system, sans-serif; height: 100%%;">
<div style="flex: 1; background: linear-gradient(135deg, #1e293b, #334155); color: white; padding: 20px; border-radius: 12px; box-shadow: 0 4px 15px rgba(0,0,0,0.15); text-align: center; border-top: 5px solid #3498db; display: flex; flex-direction: column; justify-content: center;">
<h3 style="margin-top: 0; font-size: 1.1rem; text-transform: uppercase; letter-spacing: 1px; color: #94a3b8;">Home Arena Wins</h3>
<div style="font-size: 3rem; font-weight: 800; margin: 10px 0; color: #38bdf8;">%s <span style="font-size: 1.5rem; font-weight: 400; color: #94a3b8;">(%s wins)</span></div>
<p style="margin: 0; font-size: 0.95rem; color: #cbd5e1;">Home teams winning on their own court</p>
</div>
<div style="flex: 1; background: linear-gradient(135deg, #1e293b, #334155); color: white; padding: 20px; border-radius: 12px; box-shadow: 0 4px 15px rgba(0,0,0,0.15); text-align: center; border-top: 5px solid #2ecc71; display: flex; flex-direction: column; justify-content: center;">
<h3 style="margin-top: 0; font-size: 1.1rem; text-transform: uppercase; letter-spacing: 1px; color: #94a3b8;">Away Arena Wins</h3>
<div style="font-size: 3rem; font-weight: 800; margin: 10px 0; color: #4ade80;">%s <span style="font-size: 1.5rem; font-weight: 400; color: #94a3b8;">(%s wins)</span></div>
<p style="margin: 0; font-size: 0.95rem; color: #cbd5e1;">Visiting teams winning on the road</p>
</div>
</div>
', 
sprintf("%.1f%%", home_pct), format(home_wins, big.mark=","),
sprintf("%.1f%%", away_pct), format(away_wins, big.mark=","))

cat(html_cards)
```

</div>
<div style="flex: 1.2; min-width: 320px; display: flex; flex-direction: column; justify-content: center;">

```{r render-donut-chart, fig.align='center', fig.width=6, fig.height=4}
# Prepare data for donut chart
df_chart <- data.frame(
  category = c("Home Arena", "Away Arena"),
  count = c(home_wins, away_wins)
) %>%
  mutate(
    percentage = count / sum(count) * 100,
    label = paste0(category, "\n", format(count, big.mark = ","), " (", round(percentage, 1), "%)")
  )

# Draw an interactive donut chart with hover information
plot_ly(
  df_chart,
  labels = ~category,
  values = ~count,
  type = "pie",
  hole = 0.55,
  textinfo = "none",
  hoverinfo = "text",
  text = ~paste0(category, ": ", format(count, big.mark = ","), " (", round(percentage, 1), "%)"),
  marker = list(colors = c("#3498db", "#2ecc71"), line = list(color = "#ffffff", width = 2))
) %>%
  layout(
    showlegend = TRUE,
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.1),
    margin = list(l = 10, r = 10, b = 10, t = 10)
  )
```

</div>
</div>

## Basketball Scorigami

A **Scorigami** is a visualization concept (originally popularized by Jon Bois) representing every unique final score combination that has ever occurred in a sport's history. 

In basketball, the home team's score is plotted on the horizontal axis (X-axis) and the visiting team's score is plotted on the vertical axis (Y-axis). Home team wins populate the lower-right triangle (where home score is greater than visiting score), while visiting team wins populate the upper-left triangle.

The interactive heatmap below displays all unique score combinations present in this dataset (outlier games with scores under 50 are filtered out). The color gradient indicates the frequency of each final score on a logarithmic scale.

```{r scorigami-chart}
# Filter out low-score anomalies and count home/away score combinations
scorigami_data <- games %>%
  filter(!is.na(homeScore), !is.na(awayScore)) %>%
  filter(homeScore >= 50, awayScore >= 50) %>%
  count(homeScore, awayScore)

# Create a tile plot using ggplot2
scorigami_plot <- ggplot(scorigami_data, aes(
  x = homeScore, 
  y = awayScore, 
  fill = n, 
  text = paste0("Score (Home - Away): ", homeScore, " - ", awayScore, "\nGames: ", format(n, big.mark = ","))
)) +
  geom_tile(width = 0.9, height = 0.9) +
  scale_fill_viridis_c(
    option = "viridis", 
    trans = "log10", 
    name = "Games Count"
  ) +
  labs(
    x = "Home Team Score",
    y = "Visiting Team Score"
  ) +
  theme_minimal(base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    plot.title = element_blank(),
    legend.position = "right"
  )

# Convert ggplot to interactive plotly object
ggplotly(scorigami_plot, tooltip = "text") %>%
  layout(
    margin = list(t = 20, r = 10, b = 20, l = 10)
  )
```

## Game Margin Distribution

The interactive bar chart below displays the distribution of final score margins (the absolute difference between the home and visiting team scores) across all matches. 

This visualization allows you to see the frequency of matches based on their final margin, highlighting the rarity of blowout games vs. the high occurrence of close finishes.

```{r score-difference-chart}
# Compute absolute score differences and counts
clean_games <- games %>%
  filter(!is.na(homeScore), !is.na(awayScore)) %>%
  filter(homeScore >= 50, awayScore >= 50) %>%
  mutate(score_diff = abs(homeScore - awayScore))

mean_diff <- mean(clean_games$score_diff)
median_diff <- median(clean_games$score_diff)

diff_data <- clean_games %>%
  count(score_diff)

# Create an interactive bar chart using plotly
plot_ly(
  diff_data,
  x = ~score_diff,
  y = ~n,
  type = "bar",
  marker = list(color = "#3498db", line = list(color = "#2980b9", width = 0.5)),
  hoverinfo = "text",
  text = ~paste0("Score Difference: ", score_diff, " pts\nMatches: ", format(n, big.mark = ","))
) %>%
  layout(
    xaxis = list(title = "Absolute Score Difference (Points)"),
    yaxis = list(title = "Number of Matches"),
    margin = list(t = 20, r = 10, b = 20, l = 10),
    shapes = list(
      list(
        type = "line",
        x0 = mean_diff, x1 = mean_diff,
        y0 = 0, y1 = 1,
        yref = "paper",
        line = list(color = "#e74c3c", width = 2, dash = "dash")
      ),
      list(
        type = "line",
        x0 = median_diff, x1 = median_diff,
        y0 = 0, y1 = 1,
        yref = "paper",
        line = list(color = "#e74c3c", width = 2, dash = "dot")
      )
    ),
    annotations = list(
      list(
        x = mean_diff,
        y = 0.95,
        yref = "paper",
        text = paste0("<b>Mean:</b> ", round(mean_diff, 1), " pts"),
        showarrow = FALSE,
        xanchor = "left",
        font = list(color = "#e74c3c")
      ),
      list(
        x = median_diff,
        y = 0.90,
        yref = "paper",
        text = paste0("<b>Median:</b> ", round(median_diff, 1), " pts"),
        showarrow = FALSE,
        xanchor = "right",
        font = list(color = "#e74c3c")
      )
    )
  )
```

## Relative Score Difference Margin

The interactive histogram below displays the distribution of the relative score difference across all matches, calculated as:

$$\text{Relative Difference} = \frac{|Home\ Score - Visiting\ Score|}{Home\ Score + Visiting\ Score}$$

An appropriate bin size of **0.5%** (`0.005`) was selected using the Freedman-Diaconis rule. This bin size reveals the smooth, right-skewed normal-like distribution of game competitiveness in professional basketball, where the vast majority of games conclude with a relative margin of under 10%.

```{r relative-score-difference-histogram}
# Compute relative score differences
rel_diff_data <- games %>%
  filter(!is.na(homeScore), !is.na(awayScore)) %>%
  filter(homeScore >= 50, awayScore >= 50) %>%
  mutate(rel_diff = abs(homeScore - awayScore) / (homeScore + awayScore))

mean_rel <- mean(rel_diff_data$rel_diff)
median_rel <- median(rel_diff_data$rel_diff)

# Draw plotly histogram with 0.5% (0.005) bins
plot_ly(
  rel_diff_data,
  x = ~rel_diff,
  type = "histogram",
  xbins = list(start = 0, end = 0.40, size = 0.005),
  marker = list(color = "#3498db", line = list(color = "#ffffff", width = 0.5)),
  hovertemplate = "Relative Difference: %{x}<br>Matches: %{y}<extra></extra>"
) %>%
  layout(
    xaxis = list(title = "Relative Score Difference (Absolute Difference / Total Score)", tickformat = ".1%"),
    yaxis = list(title = "Number of Matches"),
    margin = list(t = 20, r = 10, b = 20, l = 10),
    shapes = list(
      list(
        type = "line",
        x0 = mean_rel, x1 = mean_rel,
        y0 = 0, y1 = 1,
        yref = "paper",
        line = list(color = "#e74c3c", width = 2, dash = "dash")
      ),
      list(
        type = "line",
        x0 = median_rel, x1 = median_rel,
        y0 = 0, y1 = 1,
        yref = "paper",
        line = list(color = "#e74c3c", width = 2, dash = "dot")
      )
    ),
    annotations = list(
      list(
        x = mean_rel,
        y = 0.95,
        yref = "paper",
        text = paste0("<b>Mean:</b> ", round(mean_rel * 100, 2), "%"),
        showarrow = FALSE,
        xanchor = "left",
        font = list(color = "#e74c3c")
      ),
      list(
        x = median_rel,
        y = 0.90,
        yref = "paper",
        text = paste0("<b>Median:</b> ", round(median_rel * 100, 2), "%"),
        showarrow = FALSE,
        xanchor = "right",
        font = list(color = "#e74c3c")
      )
    )
  )
```

